---
title: "lab 4 - Differential Expression"
author: "ks"
date: "6/5/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 4 - Differential Expression

```{r dataset}
#source("http://bioconductor.org/biocLite.R")
library(limma)

datadir=c("data/JBC 2012")
load(file.path(datadir,"jbcComBatCor.rda"))
names(jbcComBatCor)
```

Create a plot of fold-change vs average (log) intensity

```{r MAplot}
Index1=which(jbcComBatCor$target$hour=="0h")
Index2=which(jbcComBatCor$target$hour=="16h")

d=rowMeans(jbcComBatCor$E[,Index2])-rowMeans(jbcComBatCor$E[,Index1])
a=rowMeans(jbcComBatCor$E)
par(mfrow=c(1,2))
plot(a,d,xlab="Average (log) intensity",ylab="Average log fold-change",main="MA-plot",pch=".")
smoothScatter(a,d,nrpoints=500,
              xlab="Average (log) intensity",
              ylab="Average log fold-change",
              main="MA plot")
abline(-1,0,col=7,lwd=3)
abline(1,0,col=7,lwd=3)
abline(0,0,col=2,lwd=3)
```

Why might one prefer the figure on the right?

Now let's run some t-tests for each gene.
```{r ttests}
library("genefilter")
tt=rowttests(jbcComBatCor$E,factor(jbcComBatCor$targets$hour))
tt[1,]
```

These are equivalent to the Ftest with equal variances: 

```{r Ftests}
Ft=rowFtests(jbcComBatCor$E,factor(jbcComBatCor$target$hour),var.equal=TRUE)
Ft[1,]
```

Now let's display the results using a Volcano Plot of -log10 pvalue vs (log) fold-change
```{r volcanoPlot.ssc,echo=FALSE}
tt=rowttests(jbcComBatCor$E,factor(jbcComBatCor$targets$hour))
lodt=-log10(tt$p.value)
plot(d,lodt,main="Volcano plot for t-test",xlab="Average fold-change",
	ylab="-log10 pvalue",pch=".")
# make larger data points when -log10 pvalue > 10 
#  pch = 20 gives solid dots, col=4 is blue
points(d[lodt>10],lodt[lodt>10],pch=20,col=4)
```

We can give emphasis to highly sigificant pvalues (-log10p > 10).

And again, smooth scatter can be used.

```{r smoothedVolcano,echo=FALSE}
smoothScatter(d,lodt,nrpoints=500,
              xlab="Average fold-change",
              ylab="-log10 pvalue",
              main="Volcano plot for t-test")
# If you want to highlight high lodt & fold change
# col= 6 is pink
points(d[lodt>10 & abs(d)>1],lodt[lodt>10 & abs(d)>1],pch=20,col=6)
```


Now let's see how the results change if we run a moderated t-test.
```{r modT}
design=model.matrix(~factor(jbcComBatCor$targets$hour))
fit=lmFit(jbcComBatCor$E,design)
efit=eBayes(fit)
lodmt=-log10(efit$p.value[,2])
smoothScatter(d,lodmt,nrpoints=500,
              main="Volcano plot for moderated t-test",xlab="Average fold-change",
	ylab="-log10 pvalue")
```

lab4 Supplemental code.R will show how the moderated t-test and pvalue are computed.

Let's compare the pvalues from student's t-test and the moderated t-test.
```{r ttvmodt}
par(mar=c(5,5,3,2))
plot(lodt,lodmt,pch=".",cex.axis=1.2,cex.lab=1.2,
	xlab="-log10(p) from 2-sample t-test",
	ylab="-log10(p) from moderated t-test (limma)")
abline(0,1,col=2,lwd=3)
```



```{r TopT}
topTable(efit,coef=2,n=8)
?topTable
```

Now let's look for treatment effects at time 16 hours.

```{r Trt@16}
time=jbcComBatCor$targets$hour
trts=factor(jbcComBatCor$targets$treatments, 
            levels=c("siNS","siCBP","sip300"))
design=model.matrix(~trts)
design[1:4,]
fit=lmFit(jbcComBatCor$E,design,subset=c(time=="16h"))

contr.matrix=cbind(siCBP=c(0,1,0),sip300=c(0,0,1))
contr.matrix
fitgpd=contrasts.fit(fit,contr.matrix)
fitgpd=eBayes(fitgpd)
topTable(fitgpd,n=10)
```

```{r Venn}
results=decideTests(fitgpd,adjust.method="none")
vennDiagram(results)
```

Here is the table of counts:
```{r results}
table(results[,c("sip300")],results[,c("siCBP")])
```

Both significant: 1095 + 1277 + 167 + 109 = 2648

Only siCBP significant = 1494 + 1239 = 2733

Now we can also break these down to include the information on direction of effect, up-regulation or down-regulation.
```{r VennUpDown}
vennDiagram(results,include=c("up","down"),counts.col=c("red", "blue"))
```

The counts for up-regulated genes only:
```{r countsUp}
vennCounts(results,include=c("up"))
```

Down-regulated genes only:
```{r countsDown}
vennCounts(results,include=c("down"))
```

If we want to study all pairwise comparisons, we can do this by using a different design matrix and contrast matrix.

```{r Trt@162}
design=model.matrix(~trts-1)
design[1:4,]
fit=lmFit(jbcComBatCor$E,design,subset=c(time=="16h"))

contr.matrix=cbind('siCBP-siNS'=c(-1,1,0),
                   'sip300-siNS'=c(-1,0,1),
                   'siCBP-sip300'=c(0,1,-1))
contr.matrix
fitgpd=contrasts.fit(fit,contr.matrix)
fitgpd=eBayes(fitgpd)
topTable(fitgpd,n=10)
```

These are the same differences in treatment means as in the earlier model.

Now look for genes associated with Time, Co-Reg or Time x Co-Reg interaction.
```{r mvm}
time=jbcComBatCor$targets$hour
trts=factor(jbcComBatCor$targets$treatments, 
            levels=c("siNS","siCBP","sip300"))
design=model.matrix(~trts*time)
#design
fit=lmFit(jbcComBatCor$E,design)

contr.matrix=cbind(siCBP=c(0,1,0,0,0,0),sip300=c(0,0,1,0,0,0),
                   Time=c(0,0,0,1,0,0),siCBPxTime=c(0,0,0,0,1,0),
                   sip300xTime=c(0,0,0,0,0,1))
#contr.matrix
fitfull=contrasts.fit(fit,contr.matrix)
fitfull=eBayes(fitfull)
topTable(fitfull,n=10)
```

Let's perform a global test of Ho: no difference across any level of time x treatment vs Ha: at least 1 difference.
```{r Ftest}
genesign=which(fitfull$F.p.value<0.05)
beta.p=fitfull$p.value[genesign,]
statsign=ifelse(beta.p<0.05,1,0)
vennDiagram(statsign)
```

```{r decideTests}
results <- decideTests(fitfull,adjust.method="none") 
vennDiagram(results)
```

Note: This does not filter on Ftest pvalue < 0.05.

##  Plot some individual genes

```{r Type}
plot(1:24,jbcComBatCor$E["ILMN_1726114",order(jbcComBatCor$targets$Type)],col=jbcComBatCor$targets$Type[order(jbcComBatCor$targets$Type)], pch=15,xlab="sample",ylab="log2 E", main="ILMN_1726114")
legend(1,12,levels(jbcComBatCor$targets$Type),cex=1.2,pch=15, col=1:6)
```

This gene is differentially expressed with time, and similar for all three treatment groups.

# And now ILMN_1759676
```{r gene3}
plot(1:24,jbcComBatCor$E["ILMN_1759676",order(jbcComBatCor$targets$Type)],col=jbcComBatCor$targets$Type[order(jbcComBatCor$targets$Type)],pch=15,xlab="sample",ylab="log2 E",main="ILMN_1759676",xlim=c(1,30))
legend("topright",levels(jbcComBatCor$targets$Type),cex=1.2,pch=15,
       col=1:6)
```

```{r genesymbols}
jbcComBatCor$genes["ILMN_1759676",]
```

The time effect for p300 knockdown is smaller than for the other treatment groups. p300 knockdown has lower expression of HOXC13 at baseline (0h) than the other treatment groups (siNS, siCBP).

```{r sessionInfo}
sessionInfo()
```