---
title: "Lab 5. Multiple Testing"
author: "ks"
date: "6/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Adjusted p-values

```{r data}
example.p=c(0.0041,0.0057,0.0061,0.01,.027,.029,.039,.132,.35,.57)
```

```{r bonf}
p.adjust(example.p,method="bonferroni")
```

```{r holm}
p.adjust(example.p,method="holm")
```

```{r hochberg}
p.adjust(example.p,method="hochberg")
```

```{r BH}
p.adjust(example.p,method="BH")
```

## Q values
```{r qvlib}
if(!require("qvalue")){BiocManager::install("qvalue")} 
library('qvalue')
data(hedenfalk)
```

```{r chars}
names(hedenfalk)
length(hedenfalk$p)
```


```{r hist}
# plot distribution of 3170 pvalues
# divide X axis into 20 bins, and plot density (use probability scale)
hist(hedenfalk$p,n=20,prob=T)
abline(1,0,lty=2)
```

```{r qobj}
qobj=qvalue(hedenfalk$p)
summary(qobj)
```

```{r qobj2}
plot(qobj)
```

# Sample Size calculation for differential expression 

How many samples do we need if we want to test  differential expression of a single gene in two groups? Functions to compute sample size and power when designing a study can be found in the library pwr.
```{r libs}
if(!require("pwr")){BiocManager::install("pwr")} 
library(pwr)
```

Any one value of sample size (n), effect size (d), or power, can be computed when the remaining options are specified.

pwr.t.test(n = NULL, d = NULL, sig.level = 0.05, power = NULL, type = c("two.sample", "one.sample", "paired"), alternative = c("two.sided", "less","greater"))  
e.g.
```{r compute}
pwr.t.test(n = 34, d = NULL, sig.level = 0.05, power = 0.8, 
           type = c("one.sample"), alternative = c("two.sided"))
```


Under certain simplifying assumptions, we can make an approximate sample size calculation when analyzing many genes (see Jung, 2005). For this we need to additionally specify the number of features tested and the proportion of true null features. Then, for a given FDR we can solve for an 'effective' significance level and substitute this significance level into standard power analysis software.

Here is an example. To achieve FDR=5%, solve for the sig.level under the following conditions:

   G =  4000 candidate gene features  
   pio = 0.99  Proportion of true null features
   Go = 3960    (= floor(G*pio))  No. true null features  
   G1 =   40    (= G - Go)           
   r1 = 24  true discoveries    
   power = 24/40 = 60%

```{r alphastar}
alpha.star = function(f=0.05,pio=0.99,G=4000,power=0.6){
  Go=floor(G*pio)
  r1=floor((G-Go)*power)
  eff.sign.level= r1*f/Go/(1-f)
  eff.sign.level
}
alpha.star(f=.01,power=0.6)
```

Let's apply formula from Jung (2005).
```{r ex}
a=alpha.star(f=.01,pio=0.99,G=4000,power=0.6)
d=1
2*(qnorm(1-a/2)+qnorm(.6))^2/(d)^2 + 0.5
```

This says we should analyze 37 samples in each group. For a more conservative alternative, let's use t-test formulation instead:
```{r pwrt}
pwr.t.test(d = d, sig.level = 0.00006112, power = 0.6, 
           type = c("two.sample"), alternative = c("two.sided"))
```

This tells us we should analyze 41 samples in each group.

```{r sI}
sessionInfo()
```
