---
title: "Gene Set Enrichment Analysis"
author: "ks"
date: "6/10/2020"
output: html_document
---

# {.tabset}

## Data

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
if (!require(genefilter)) BiocManager::install("genefilter")
if (!require(msigdbr)) install.packages("msigdbr")
library(msigdbr)
library(tidyverse)
library(genefilter)
library(limma)
```

We will continue to analyze the prostate cancer cell line data, focussing on the control samples. All cells are treated with androgen at 0 hrs. The genes that change expression between 0 and 16 hours are genes that are regulated by androgen. 

We load the data (from JBC (2012)), and subset to the control reactions (treatment = siNS).
```{r ReadData}
jbcdir=c("data/JBC 2012")
load(file.path(jbcdir,"jbcdat.rda"))

geoid <- rownames(jbcdat$targets)[jbcdat$targets$treatment == "siNS"]
subsetEobj <- function(dataobj,geonames){
  subobj <-NULL
     subobj$targets <- dataobj$targets[geonames,]
     subobj$E       <- dataobj$E[,geonames]
     subobj$genes   <- dataobj$genes
     subobj
}

siNSobj <- subsetEobj(jbcdat,geoid)
# check the annotation and expression are both subset identically
identical(rownames(siNSobj$targets),colnames(siNSobj$E))
```

## Moderated t-tests 

Let's rank the genes using a moderated t-test.
```{r modT}
design <- model.matrix(~factor(siNSobj$targets$hour))
fit  <- limma::lmFit(siNSobj$E,design)
efit <- limma::eBayes(fit)
```

Top hits:

```{r TopT}
topT <- topTable(efit,coef=2,n=10)
topT
```

Volcano plot:

```{r volcanoplot, echo=FALSE}
volcanoplot(efit, coef = 2, style = "p-value", highlight = 10, names = siNSobj$genes$Symbol, hl.col="blue",
            xlab = "Log2 Fold Change", ylab = NULL, pch=16, cex=0.35)
```

## Feature Annotation

What information is provided on the features in our top hits?

```{r genenames}
siNSobj$genes[rownames(topT),]
```

Notice that 2 of the top 10 hits are for the same Entrez Gene ID (6446). And they have different associated Symbols (that's problematic!) If we were analyzing these data, it'd be worth looking for updated feature annotation for this array. 

Let's investigate the number of probes/gene. I'll do this by the Entrez Gene ID. Here are the total number of features:
```{r nfeatures}
length(siNSobj$genes$Entrez_Gene_ID)
```

Here is the number of unique Entrez Gene IDs:
```{r ugid}
length(unique(siNSobj$genes$Entrez_Gene_ID))
```

There are 47231 features, but fewer Entrez Gene IDs.

Let's tabulate the number of features for the Entrez Gene IDs. Then we'll compute a table of the number of features/gene.
```{r chars}
ts <- table(siNSobj$genes$Entrez_Gene_ID)
table(ts)
```

So 21592 Entrez Gene IDs have just 1 feature, but 5809 have 2 features, 2127 have 3 features, etc. It turns out 3270 features are captured by a missing Entrez Gene ID.

Let's check how the values of the Entrez Gene IDs again.
```{r arrayannot}
head(siNSobj$genes)
```

The Entrez Gene IDs are stored as a character value. To find the features with missing IDs, I'll cast the Entrez Gene ID as numeric and then identify the values that are not positive numbers. 

```{r find3270}
idx <- which(as.numeric(siNSobj$genes$Entrez_Gene_ID)>0)
idxc <- c(1:nrow(siNSobj$genes))[-idx]
table(siNSobj$genes$Entrez_Gene_ID[idxc],useNA="always")
```

This shows 3270 Entez_Gene_IDs are missing and 1 is coded as missing (NA). Let's see the Entrez_Gene_ID for the first 3 missing ids.

```{r blanks}
siNSobj$genes$Entrez_Gene_ID[idxc][1:3]
```

It is good that we identifed the features without Entrez Gene IDs.  Later, when we do the gene set enrichment analysis, we will restrict our analysis to features with Entrez Gene IDs. A feature that doesn't have an ID will never get picked in a gene set, so it doesn't belong in our variable list.

## MSigDB

I will get my list of gene sets from the Molecular Signatures database (available here:  https://www.gsea-msigdb.org/gsea/msigdb/index.jsp). I don't have strong opinions on the quality of different databases. I'm selecting this one because (1) the authors of the limma package used it previously for demonstrating their package and (2) it includes well known pathway databases such as REACTOME, BIOCARTA and KEGG.  

We'll download the human gene sets.
```{r human-msigs}
m_df = msigdbr(species = "Homo sapiens")
head(m_df)
```

The sessionInfo() later will tell us this is version 7.0.1, which is not up to date with the website which is up to version 7.1.

Here are the names of the 8 major gene set collections.
```{r tabsets}
unique(m_df$gs_cat)
```

We'll need to organize one of these collections into a list of lists, with each element representing a gene set. I'm going to consider the curated gene sets in collection C2.
```{r c2}
hs.c2.msig <- m_df %>% 
                filter(gs_cat=="C2")
hs.c2.gsnames  <- unique(hs.c2.msig$gs_name)
length(hs.c2.gsnames)
```

There are a total of 5501 curated gene sets. (The website is up to 5529).

Now I'll make the list of gene sets. I welcome advice on an easier way to program this.
```{r newmsigdb}
nsets <- length(hs.c2.gsnames)
hs.c2.msList <- list()
for (i in 1:nsets){
    idx <- which(hs.c2.gsnames[i]==hs.c2.msig$gs_name)
    hs.c2.msList[i] <- list(hs.c2.msig$entrez_gene[idx])
}
names(hs.c2.msList) <- hs.c2.gsnames
```  
  
Here is the distribution of gene set sizes.
```{r gssizes}
summary(map_int(hs.c2.msList,length))
```  

The median gene set size is 34, and the smallest is only 5.

Here are the first 2 sets.
```{r sets}
  hs.c2.msList[1:2]
```

## GSEA

To test for gene set enrichment, we need to map the array features to the lists of Entrez Gene IDs. First, we'll restrict our dataset to the subset of array features that have an ID.

```{r decide}
gidna <- ifelse(siNSobj$genes$Entrez_Gene_ID=="",NA,siNSobj$genes$Entrez_Gene_ID)
idx <- which(!is.na(gidna))
c2.indices <- ids2indices(hs.c2.msList,siNSobj$genes$Entrez_Gene_ID[idx])
```

Then, we provide the entire data set with the design matrix specifying the two-group comparison.
```{r camera}
design <- model.matrix(~factor(siNSobj$targets$hour))

cam.16h <- camera(siNSobj$E[idx,],c2.indices,design,
                  inter.gene.cor=0.01)
head(cam.16h)
```

Nice! Many of these gene sets are known for their response to androgen.

Let's check the number of genes in the top gene set.
```{r gs}
    length(unlist(hs.c2.msList["NELSON_RESPONSE_TO_ANDROGEN_UP"]))
```

And how about the number of features representing these genes...
```{r gsi}
    length(unlist(c2.indices["NELSON_RESPONSE_TO_ANDROGEN_UP"]))
```

This gene set has 84 genes, but there are $>$ 1.5 features per gene on average. 

The naming of the top 2 gene sets makes me wonder how many of the same genes appear in the top 2 sets?
```{r overlap}
    intersect(unlist(hs.c2.msList["NELSON_RESPONSE_TO_ANDROGEN_UP"]),
              unlist(hs.c2.msList["WANG_RESPONSE_TO_ANDROGEN_UP"]))
```
There are 8.This represents 29\% of the second list. 
```{r gsw}
    length(unlist(hs.c2.msList["WANG_RESPONSE_TO_ANDROGEN_UP"]))
```

Also, the label in the table that says "NGenes" really means the number of features. 
Having multiple features for some of the genes makes this an interesting gene set to explore the effect of correlation on gene set testing.  Let's try this by feeding the function a list of t-tests.

```{r decidet}
cam.16h <- cameraPR(efit$t[idx,2],c2.indices)
head(cam.16h)
```

The top 3 gene sets are the same, but with MUCH more significant pvalues. The significance is inflated because the method cannot adjust for correlation between features when supplied with the t-tests. 

Let's look at a figure of the distribution of these test statistics:
```{r densityt}
plot(density(efit$t[idx,2]),main="",xlab="t-statistic")
rug(efit$t[idx,2][c2.indices$NELSON_RESPONSE_TO_ANDROGEN_UP],col=2)
lines(density(efit$t[idx,2][c2.indices$NELSON_RESPONSE_TO_ANDROGEN_UP]),col=2)
legend(5,.3,col=c(1,2),lty=c(1,1),c("All t-stats","gene set t-stats"),cex=0.7)
```




## REACTOME pathways

REACTOME is one pathway database contributing to the curated gene sets. We can test just these pathways by subsetting the gene set list.

```{r reactome-pathway-number}
rt=which(substr(names(hs.c2.msList),1,8)=="REACTOME")
length(rt)
```

This represents 27\% of the gene sets in our list.

```{r reactome}
gsea.rt <- camera(siNSobj$E[idx,],c2.indices[rt],design,
                  inter.gene.cor=0.01)
head(gsea.rt)
```

Warning! Gene set tests typically suffer from multiple testing. We rarely test just 1 hypothesis, but test many gene sets.  We will study multiple testing next week.

## SessionInfo

```{r sessionInfo}
sessionInfo()
```