---
title: "Estimating FDR with qvalues"
author: "ks"
date: "6/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require("qvalue")){BiocManager::install("qvalue")}
```

## p-values

We load the list of 3,701 p-values from Storey & Tibshirani's analysis of the Hedenfalk et al. data that are provided in the qvalue package. 

```{r estfdr}
library(qvalue)
data(hedenfalk)
names(hedenfalk)
```


```{r pvalue}
length(hedenfalk$p)
```

There were 3,170 pvalues. Let's plot them in a histogram.

```{r histp}
hist(hedenfalk$p,probability=TRUE)
abline(1,0,lty=2)
```

The histogram of the p-values shows an excess of small p-values, with the rectangular shape of a uniform distribution for p>0.5.

How many p-values are < 0.05?

```{r p0.05}
sum(hedenfalk$p < 0.05)
```

Using a regular p<0.05 cutoff, we'd find 605 genes differentially expressed. We know that because of multiple testing, we'd expect 3,107 x 5\% = 159 significant tests (26% of 605). That's a high error rate. Let's control the false-discovery rate at 5\% using the Benjamini and Hochberg adjusted p-value.

```{r bhadj}
bhadjp <- p.adjust(hedenfalk$p,method="BH")
sum(bhadjp < 0.05)
```

Using a 5\% false-discovery rate cutoff, we'd find 94 genes significant (Benjamini & Hochberg adjusted p < 0.05).   

## q-values

We compute the q-values from the list of pvalues, and create several figures summarizing the results.

```{r qvalue}
qv <- qvalue(hedenfalk$p)
plot(qv)
```

The analysis results are summarized in 4 figures. The top left fits a cubic spline (smooth curve) to the estimates of $\hat{\pi}_0 (\lambda)$. The dots become more variable as $\lambda$ approaches 1, but the estimate $\hat{\pi}_0 (\lambda)$ from the curve is more stable.  This curve tells us $\hat{\pi}_0 (1)$ = 0.67.

The top right figure is the # significant tests as a function of the q-value cutoff (estimate of FDR). The large steps for the q-values <0.025 show cutoffs for which a number of tests become significant.  

The bottom left figure tells us the q-value for each p-value cutoff (0<p<0.015). If you rejected your tests with unadjusted p<0.015, you would expect 10% of the tests to be false.

The final figure (bottom-right) gives the expected number of false-positive results given the number of top hits (0-300). If you reject 300 tests, you would expect fewer than 30 errors (false-discovery rate < 10%). 

The summary function tabulates the results to count significant genes at a series of standard cut-offs.

```{r summaryqvalue}
summary(qv)
```

This gives the estimate that 67\% (pi0) of the genes are not differentially expressed.  The row named p-value is the count for the raw (unadjusted) pvalues. We see 605 pvalues are less than 0.05, the same number obtained above.  A q-value cutoff of 5% yields 162 significant tests.

We can also compute the number of significant tests directly, using the following command.
```{r qvals}
sum(qv$qvalues <0.05)
```

Since the q-value = BH-adjusted p x $\hat{\pi}_0$, we can check this result. 
```{r computeqvals}
sum( bhadjp * qv$pi0  <0.05)
```

Yes, it's the same! 

```{r sessioninfo}
sessionInfo()
```
