---
title: "2.b. Step-by-step Gene Network construction-mouse liver"
author: "ks"
date: "7/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Tutorials for WGCNA package in R from the following website

http://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/index.html

Some omissions were made, but mostly this is taken verbatim from the tutorials.


```{r loadlibs}
library(cluster)
library(Hmisc)	# install it for the C-index 
library(MASS)  # standard, no need to install
library(class)  # standard, no need to install
library(WGCNA)
library(flashClust)
```

# 2. Network Construction and Module Detection

##2a. Automatic construction of gene network and module identification 

(see w10-genenetworks-mouseliver.Rmd)

## 2b. Network construction (step-by-step) and module detection

First, we allow multi-threading within WGCNA if your computer has this ability. This helps speed up certain calculations. At present this call is not necessary for the code to work. Any error here may be ignored but you may want to update WGCNA if you see one.

```{r multithread}
allowWGCNAThreads();
```

This loads the data we created in: w10-genenetworks-mouseliver.Rmd.

```{r loadData}
# Load the data saved in the first part
lnames <- load(file = "FemaleLiver-01-dataInput.RData");
#The variable lnames contains the names of loaded variables.
lnames
```

### 2b.1 Choosing the soft-thresholding power: analysis of network topology

Constructing a weighted gene network entails the choice of the soft 
thresholding power β to which co-expression
similarity is raised to calculate adjacency [1]. The authors of [1] 
have proposed to choose the soft thresholding power
based on the criterion of approximate scale-free topology. We refer the 
reader to that work for more details; here
we illustrate the use of the function pickSoftThreshold that performs 
the analysis of network topology and aids the
user in choosing a proper soft-thresholding power. The user chooses a 
set of candidate powers (the function provides
suitable default values), and the function returns a set of network 
indices that should be inspected, for example as follows:
```{r connect}
# Choose a set of soft-thresholding powers
powers <- c(c(1:10), seq(from = 12, to=20, by=2))
# Call the network topology analysis function
sft <- pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
# Plot the results:
#sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
    xlab="Soft Threshold (power)",
    ylab="Scale Free Topology Model Fit,signed R^2",type="n",
    main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
    labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
      xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
      main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```

The result is shown in the Fig. above. We choose the power 6, which is the lowest power for which the scale-free topology fit
index curve flattens out upon reaching a high value (in this case, roughly 0.90).


###2b.2 Co-expression similarity and adjacency

We now calculate the adjacencies, using the soft thresholding power 6:
```{r adjacency}
softPower <- 6;
adjacency <- adjacency(datExpr, power = softPower);
```

###2b.3 Topological Overlap Matrix (TOM)

To minimize effects of noise and spurious association:
```{r TOM}
TOM <- TOMsimilarity(adjacency);
dissTOM <- 1-TOM
```

###2b.4 Clustering using TOM

We now use hierarchical clustering to produce a hierarchical 
clustering tree (dendrogram) of genes. Note that we use
the function hclust that provides a much faster hierarchical 
clustering routine than the standard hclust function.
```{r genetree}
# Call the hierarchical clustering function
geneTree <- flashClust(as.dist(dissTOM), method = "average");
# Plot the resulting clustering tree (dendrogram)
#sizeGrWindow(12,9)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
      labels = FALSE, hang = 0.04);
```

Branches of the dendrogram group together densely
interconnected, highly co-expressed genes. Module identification amounts to the identification of individual branches
(”cutting the branches off the dendrogram”).

```{r cuttree}
# We like large modules, so we set the minimum module size relatively high:
minModuleSize <- 30;
# Module identification using dynamic tree cut:
dynamicMods <- cutreeDynamic(dendro = geneTree, distM = dissTOM,
                      deepSplit = 2, pamRespectsDendro = FALSE,
                      minClusterSize = minModuleSize);
table(dynamicMods)
```

The function returned 22 modules labeled 1–22 largest to smallest. Label 0 is reserved for unassigned genes. The
above command lists the sizes of the modules. We now plot the module assignment under the gene dendrogram:
```{r plottree}
# Convert numeric lables into colors
dynamicColors <- labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
#sizeGrWindow(8,6)
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                  dendroLabels = FALSE, hang = 0.03,
                  addGuide = TRUE, guideHang = 0.05,
                  main = "Gene dendrogram and module colors")
```

###2b.5 Merging of modules whose expression profiles are very similar

The Dynamic Tree Cut may identify modules whose expression profiles are very similar. It may be prudent to merge
such modules since their genes are highly co-expressed. To quantify co-expression similarity of entire modules, we
calculate their eigengenes and cluster them on their correlation:
```{r Eiggenes}
# Calculate eigengenes
MEList <- moduleEigengenes(datExpr, colors = dynamicColors)
MEs <- MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss <- 1-cor(MEs);
# Cluster module eigengenes
METree <- flashClust(as.dist(MEDiss), method = "average");
# Plot the result
#sizeGrWindow(7, 6)
plot(METree, main = "Clustering of module eigengenes",
        xlab = "", sub = "")
MEDissThres <- 0.25
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")
```

We choose a height cut of 0.25, corresponding to correlation of 0.75, to merge (see Fig.):
```{r mergecut}
# Call an automatic merging function
merge <- mergeCloseModules(datExpr, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# The merged module colors
mergedColors <- merge$colors;
# Eigengenes of the new merged modules:
mergedMEs <- merge$newMEs;
```

To see what the merging did to our module colors, we plot the gene dendrogram again, with the original and merged
module colors underneath (see Figure).

```{r plotafermerge}
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
          c("Dynamic Tree Cut", "Merged dynamic"),
          dendroLabels = FALSE, hang = 0.03,
          addGuide = TRUE, guideHang = 0.05)
```

Prep merged data and save for later use.
```{r cleanandsave}
# Rename to moduleColors
moduleColors <- mergedColors
# Construct numerical labels corresponding to the colors
colorOrder <- c("grey", standardColors(50));
moduleLabels <- match(moduleColors, colorOrder)-1;
MEs <- mergedMEs;
# Save module colors and labels for use in subsequent parts
save(MEs, moduleLabels, moduleColors, geneTree, file = "FemaleLiver-02-networkConstruction-stepByStep.RData")
```

```{r sI}
sessionInfo()
```