---
title: "Example CCA from PMA package"
author: "Ks"
date: "7/29/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Sparse canonical correlation analysis (CCA) is included in the PMA library we used in week 2 of this class for sparse PCA.

```{r library}
#install.packages("PMA")
library(PMA)
```

# code is from Reference Manual PMA.pdf
https://cran.r-project.org/web/packages/PMA/PMA.pdf


## Simulated data example

How do we simulate data, to illustrate sparse CCA? Here's how Witten and Tibshirani do it:

```{r simdat}
set.seed(3189)
u <- matrix(c(rep(1,25),rep(0,75)),ncol=1)
v1 <- matrix(c(rep(1,50),rep(0,450)),ncol=1)
v2 <- matrix(c(rep(0,50),rep(1,50),rep(0,900)),ncol=1)
x <- u%*%t(v1) + matrix(rnorm(100*500),ncol=500)
z <- u%*%t(v2) + matrix(rnorm(100*1000),ncol=1000)
```


We can run CCA with default settings and request 3 components.
```{r cca}
out <- CCA(x,z,typex="standard",typez="standard",K=3)
print(out, verbose = TRUE) 
```



Or can use CCA.permute to choose optimal parameter values.

```{r permout}
perm.out <- CCA.permute(x,z,typex="standard",typez="standard",nperms=7)
print(perm.out)
```

```{r plotperm.out}
plot(perm.out)
```

Now we can fit a model with the best shrinkage parameters.

```{r bestpenalty}
out <- CCA(x,z,typex="standard",typez="standard",K=1,penaltyx=perm.out$bestpenaltyx,penaltyz=perm.out$bestpenaltyz, v=perm.out$v.init)
print(out)
```


```{r si}
sessionInfo()
```