---
title: "Analysis of The Cancer Genome Atlas (TCGA) data"
author: "ks"
date: "July 29, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Colon Adenocarcinoma (COAD) data  

The DNA methylation platform changed during the time the TCGA samples were begin analyzed. The earlier samples were analyzed using the HumanMethylation27 platform. Today we're 2 generations past that one (HumanMethylation450 and MethylationEPIC).

I've selected for analysis the samples with HumanMethylation27 data since the tumors measured using this platform have the most complete annotation (e.g. tumor site, CIMP-designation, tumor stage). I previously downloaded these data along with patient clinical information from the package: curatedTCGAData. I got the RNAseq data from GEO: GSE62944. 
```{r LOADanalysisdataset}
dataDir = c("pm579data/COAD")
load(file.path(dataDir,"coadhm27-AnalysisSet.rda"))
load(file.path(dataDir,"coadhm27-gex-AnalysisSet.rda"))
load(file.path(dataDir,"coadhm27-clindat-AnalysisSet.rda"))
```

Outline an anlaysis you'd like to see of this data, and begin to explore it.



```{r SessionInfo}
sessionInfo()
```