---
title: "Simulation in R"
author: "ks"
date: "5/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(ggplot2)) {install.packages("ggplot2")}
library(ggplot2)
```

We are going to simulate the following scenario:

In a population of 10,000 young women, the average sbp in women not carrying the gene is 120 mm Hg with a sd of 11. In the 1\% of the population that carries the gene, the average blood pressure is 131 (1 sd above the mean).
Simulate a population of 10,000 young women.

If you sampled 500 of the women at random, how many gene carriers do you expect in your sample?

If you measured sbp in all 10,000 young women and selected the 500 with the highest blood pressure, how many gene carriers would you expect now? The same number? More? Fewer? Do you need more information?

**Simulate sbp for 10,000 young women and plot the sbp distributions in gene carriers and non-carriers.**
```{r simulate}
#set.seed(200)
df <- data.frame(
        carrier=factor(c(rep("No",9900),rep("Yes",100))),
            sbp=round(rnorm(10000, mean=120, sd=11),2)
      )
df$sbp[9901:10000] <- df$sbp[9901:10000] + 11
```

```{r plot}
ggplot(df, aes(x=sbp, fill=carrier)) +
  geom_density()

t.test(sbp~carrier,data=df)
```

**Sample 500 people at random. How many are gene carriers?**

```{r rsamp}
idx <- sample(1:10000,500)
sum(df$carrier[idx]=="Yes")
```

**Now go rank the women on SBP and select the 500 with the higest values. Now how many are gene carriers?**

```{r topsbp}
rsbp <- rank(-df$sbp)
idx <-  which(rsbp <= 500)

sum(df$carrier[idx]=="Yes")
```

```{r sessioninfo}
sessionInfo()
```
