---
title: "Lung Cancer in AA & EA"
author: "ks"
date: "6/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadlibraries}
if(!require("GEOquery")) {BiocManager::install("GEOquery")}
library(GEOquery)
library(tidyverse)
library(purrr)
```

## Read GEO dataset

GEO accession: GSE102287 has both the miRNA and mRNA data sets.
GEO accession: GSE101929 has the mRNA data, let's start with downloading just that.

These data are from https://pubmed.ncbi.nlm.nih.gov/29196495/

```{r geoquery}
gse101929 <- getGEO('GSE101929',GSEMatrix=TRUE)
show(gse101929)
```

## Gene expression data

```{r gexdata}
geodat <- gse101929$`GSE101929_series_matrix.txt.gz`
luad <- NULL
luad$E <- exprs(geodat)    # exprs() accesses the gene expression values
luad$E[1:4,1:3] 
```

## Sample annotation data

```{r pdata}
pData(geodat)$title 
```

And compare the column order from the gene expression data to the row order of the sample annotation data.

```{r checkorder}
identical(colnames(luad$E),
          as.character(pData(geodat)$geo_accession))
```

Good. The data matrices are identically ordered, linking the data in each.

```{r pData}
names(pData(geodat))
```

Now let's explore what's in these .
```{r pdatavalues}
head(pData(geodat))
```

```{r targets}
char1 <- strsplit(as.character(pData(geodat)$characteristics_ch1)," ")
ptid <- map_chr(char1,pluck,3)
#cbind.data.frame(pData(geodat)$characteristics_ch1,ptid)
pdat <- pData(geodat)[,c(45:51,54:55)]
cn <- colnames(pdat)
cn <- sub(":ch1","",cn)
cn <- sub("death due to lung cancer \\(all years\\)","status",cn)
cn <- sub("survival \\(years\\)","time",cn)
colnames(pdat) <- sub("tumor_normal status","tissue",cn)
luad$targets <- cbind.data.frame(ptid,pdat)
luad$targets$pys <- ifelse(luad$targets$'smoking pack years'=="y",NA, luad$targets$'smoking pack years')
head(luad$targets)
```

All the variables are strings. Let's make them either numeric or factors.
```{r vartype}
luad$targets$age <- as.numeric(luad$targets$age)
luad$targets$status <- factor(luad$targets$status)
luad$targets$gender <- factor(luad$targets$gender)
luad$targets$race <- factor(luad$targets$race)
luad$targets$Stage <- factor(luad$targets$Stage)
luad$targets$time <- as.numeric(luad$targets$time)
luad$targets$tissue <- factor(luad$targets$tissue)
luad$targets$pys <- as.numeric(luad$targets$pys)
head(luad$targets)
```


These data represent tumor and matched non-tumor lung tissue in European Americal and African American lung cancer patients. 

## Gene annotation data

```{r fdata}
names(fData(geodat))
names(fData(geodat))[c(1,11:16)]
luad$genes <- cbind.data.frame(fData(geodat)[,c(1,11:16)])
head(luad$genes)
```


```{r fdata}
#head(luad$genes)
table(luad$genes$ENTREZ_GENE_ID=="")
```

Let's restrict the data set to non-missing Entrez Gene IDs.
```{r fdata}
table(luad$genes$ENTREZ_GENE_ID=="")
idx <- which(luad$genes$ENTREZ_GENE_ID!="")
luad$E <- luad$E[idx,]
luad$genes <- luad$genes[idx,]
dim(luad$E)
dim(luad$E)
table(luad$genes$ENTREZ_GENE_ID=="")
```

Now we can save the data set for later analysis.

```{r savedata}
    names(luad)
    save(luad,file = c("data/luad.rda"))
```

```{r sessioninfo}
sessionInfo()
```
