---
title: "Class Activity -- Filtering Significant Hits"
author: "Kim Siegmund"
date: "`r Sys.Date()`"
output: html_document
---


## 2-sample t-test: 10,000 genes 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(genefilter)
library(limma)
library(tidyverse)
```

I simulate the exact same data as in the previous example and perform a 2-sample t-test on all genes.

```{r jbcdat, echo=FALSE}
load(file = c("../data/JBC 2012/jbcdat.rda"))
geoid <- rownames(jbcdat$targets)[jbcdat$targets$treatment == "siNS"]
subsetEobj <- function(dataobj,geonames){
  subobj <-NULL
     subobj$targets <- dataobj$targets[geonames,]
     subobj$E       <- dataobj$E[,geonames]
     subobj$genes   <- dataobj$genes
     subobj
}
siNSobj <- subsetEobj(jbcdat,geoid)

design <- model.matrix(~factor(siNSobj$targets$hour))
fit  <- limma::lmFit(siNSobj$E,design)
efit <- limma::eBayes(fit)
sds <- sqrt(efit$s2.post)

# design values
nsamp <- 100       # number of samples
n1     <- nsamp/2    # equal sample sizes in each group
ng     <- 10000     # number of genes
pinull    <- 0.95      # proportion of null genes
G0     <- ng*pinull  # number of null genes
set.seed(99)
ssds <- sample(sds,ng) # sample, and randomize the order
```


```{r simdat, echo=FALSE}
simdat <- function(se,delta=0.02,nsamp=100,ng=10000,mn=0,pi0=0.95){
  n1 <- nsamp/2   
  gp  <- factor(rep(c("group1", "group2"), each=n1))
    sdat <- replicate(nsamp,rnorm(n = ng, mean = mn, sd = se))
     deg <- rep(0,ng)
     # differential expressed genes (deg) occur in first G1 rows
      G1 <- ng*(1-pi0)
  deg[1:G1] <- 1
  sdat[c(1:G1),c(1:n1)] <- sdat[c(1:G1),c(1:n1)] + delta
    
  list(E=sdat,targets=gp,genes=deg)
}
```


```{r rsimdat}
set.seed(82) 
sim1 <- simdat(ssds,delta=0.11)
tt=genefilter::rowttests(sim1$E,sim1$targets)
```


Here's the volcano plot.
```{r volcano, echo=FALSE}
d    <- tt$dm
lodt <- (-log10(tt$p.value))

plot(d,lodt,main="Volcano plot for t-test",xlab="Log2 Fold Change",
	ylab="-log10 (P-value)",pch=".")
# make larger data points when fdr adjusted p < 0.05
#  pch = 20 gives solid dots, col=4 is blue
adj.p <- p.adjust(tt$p.value,method="BH")
points(d[adj.p<0.05],lodt[adj.p<0.05],pch=20,cex=0.5,col=4)
```


What if we filter on |lfc| > 0.15 *after* we've performed our FDR multiple-test correction? Scientists want to do this all the time.  The (false) intuition is that if there is a 5% error rate in my discoveries, picking the discoveries with largest difference in means will get the largest effect size and be 'real'.  But, does the difference in means = effect size?

```{r BH_adjustedp_thenfilt}
# test all genes and add BH-adjusted p and truth about DGE to result matrix
BH.adjp <- p.adjust(tt$p.value,method="BH")
tt <- cbind.data.frame(tt,BH.adjp,altg = sim1$genes)
# subset genes to those with BH-adj p < 0.05
subtt <- tt[BH.adjp < 0.05,]
dim(subtt)
# filter, saving those with high log-foldchange
fidx <- which(abs(subtt$dm) > 0.18)
x <- table(altg = subtt$altg[fidx],
            sign = (subtt$BH.adjp<0.05)[fidx])
x
prop.table(x,2)
```
What is the false-discovery rate now?

Let's recreate the volcano plot and show the false-positives in red.
```{r volcano2, echo=FALSE}
d    <- tt$dm
lodt <- (-log10(tt$p.value))

plot(d,lodt,main="Volcano plot for t-test",xlab="Log2 Fold Change",
	ylab="-log10 (P-value)",pch=".")
# make larger data points when fdr adjusted p < 0.05
#  pch = 20 gives solid dots, col=4 is blue
adj.p <- p.adjust(tt$p.value,method="BH")
points(d[adj.p<0.05],lodt[adj.p<0.05],pch=20,cex=0.5,col=4)
points(d[adj.p<0.05 & sim1$genes==0],
       lodt[adj.p<0.05 & sim1$genes==0],pch=20,cex=0.6,col=2)
segments(0.18,0,0.18,10,lty=3)
segments(-0.18,0,-0.18,10,lty=3)
```

Notice how the false-positives were towards the bottom of the statistically significant results and how filtering on the log fold change *increased* the false discovery rate of our final filtered subset above 5\%. For this dataset it was 13\%.

This is just 1 data set. We should repeat this in a simulation study to verify that the inflated error rate is reproducible.  XXXX et al. says is is! 

### Session Information
```{r sI}
sessionInfo()
```
