---
title: "Analysis of simulated multi-block data"
author: "Ks"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Multiblock data refers to data sets with variables that can be grouped into 'blocks' of data measured on the same samples. Examples include having gene expression, DNA methylation, protein expression, and micro RNA all measured on the same samples. Each data type constitutes a 'block' of data.

Last week we anaylzed data with 2 blocks.  This week, let's extend that to data with k>2 blocks.

```{r library}
if (!require(PMA)) {install.packages("PMA"); 
library(PMA)}
if (!require(corrplot))  {install.packages("corrplot"); library(corrplot)}
```

# code is from Reference Manual PMA.pdf
https://cran.r-project.org/web/packages/PMA/PMA.pdf

## Simulated data example

Generate 3 data sets so that first 25 features are correlated across
the data sets...

```{r simdat}
set.seed(123)
u <- matrix(rnorm(50),ncol=1)
v1 <- matrix(c(rep(.5,25),rep(0,75)),ncol=1)
v2 <- matrix(c(rep(1,25),rep(0,25)),ncol=1)
v3 <- matrix(c(rep(.5,25),rep(0,175)),ncol=1)
x1 <- u%*%t(v1) + matrix(rnorm(50*100),ncol=100)
x2 <- u%*%t(v2) + matrix(rnorm(50*50),ncol=50)
x3 <- u%*%t(v3) + matrix(rnorm(50*200),ncol=200)
xlist <- list(x1, x2, x3)
```


Describe the dataset.  How many observations are there?
How many 'blocks' of data? How many variables in each block?  
```{r}


```


Describe the matrix: u%*%t(v1)
```{r}

```

Describe the matrix x1.
x1 = u%*%t(v1) + matrix(rnorm(50*100),ncol=100)


```{r}
corrplot::corrplot(x2,is.corr=FALSE,addgrid.col=NA)
```



What does matrix x1 have in common with x2? Is there correlation between variables in x1, x2, & x3?
```{r}


```





From the vignette: Run MultiCCA.permute w/o specifying values of tuning parameters to try. The function will choose the lambda for the ordered data set. Then permutations will be used to select optimal sum(abs(w)) for standard data sets.

```{r}
perm.out <- MultiCCA.permute(xlist, type=c("standard", "ordered",
"standard"))
print(perm.out)
```

```{r}
plot(perm.out)
```

```{r}
out <- MultiCCA(xlist, type=c("standard", "ordered", "standard"),
penalty=perm.out$bestpenalties, ncomponents=2, ws=perm.out$ws.init)
print(out)
```


How many variables are selected from each of the 3 blocks of data? Consider just the first cannonical variable.
```{r}

```

How many of the first 25 variables are identified?
```{r}

```



Verify out$cors[1]=1.902578 from the output.    

HINT: cor(x1cc1,x2cc1) + cor(x1cc1,x3cc1) + cor(x2cc1,x3cc1)
```{r}


```


```{r}

```




```{r si}
sessionInfo()
```