---
title: "Example CCA from PMA package"
author: "Ks"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Sparse canonical correlation analysis (CCA) is included in the PMA library.

```{r library}
if (!require(PMA)) {install.packages("PMA"); 
library(PMA)}
```

# code is from Reference Manual PMA.pdf
https://cran.r-project.org/web/packages/PMA/PMA.pdf


## Simulated data example

How do we simulate data, to illustrate sparse CCA? Here's how Witten and Tibshirani do it:

```{r simdat}
set.seed(3189)
u <- matrix(c(rep(1,25),rep(0,75)),ncol=1)
v1 <- matrix(c(rep(1,50),rep(0,450)),ncol=1)
v2 <- matrix(c(rep(0,50),rep(1,50),rep(0,900)),ncol=1)
x <- u%*%t(v1) + matrix(rnorm(100*500),ncol=500)
z <- u%*%t(v2) + matrix(rnorm(100*1000),ncol=1000)
```


We can run CCA with default settings and request 3 components.
```{r cca}
out <- CCA(x,z,typex="standard",typez="standard",K=3)
print(out, verbose = TRUE) 
```



Or can use CCA.permute to choose optimal parameter values.

```{r permout}
perm.out <- CCA.permute(x,z,typex="standard",typez="standard",nperms=7)
print(perm.out)
```

```{r plotperm.out}
plot(perm.out)
```

Now we can fit a model with the best shrinkage parameters.

```{r bestpenalty}
out <- CCA(x,z,typex="standard",typez="standard",K=1,penaltyx=perm.out$bestpenaltyx,penaltyz=perm.out$bestpenaltyz, v=perm.out$v.init)
print(out)
```


How many non-zero elements are there?
```{r coefs}
sum(out$u[,1]!=0)
sum(out$v[,1]!=0)
```

I did not feed the CCA function matrices with scaled column vectors, but the function did the scaling of the data before performing CCA. To get the canonical variables out, I will have to scale the columns of my input matrix before I multiply the rows by the vector of coefficients. By scale, I mean center the data to have mean 0, sd 1.

Let's plot the first 2 components:
```{r plotcc1s}
xc <- scale(x)
zc <- scale(z)
xcc1<- xc%*%out$u[,1]
zcc1<- zc%*%out$v[,1]
plot(xcc1,zcc1,
     main=paste("cor=",round(cor(xcc1,zcc1),7)))
```

The correlation matches the output from "print(out)".


```{r si}
sessionInfo()
```