---
title: "Homework 1"
author: "ks"
date: "due 5/23/2018"
output: html_document
---

# Homework Goals

In this homework you will learn the following: 

1. How to use R Markdown to create a report
2. How to read data from different formats into R
3. How to use GitHub to turn in the code & results (.rmd and .html files) 

Next week we will discuss, in class, our homework experience. 

###R Markdown 

R Markdown is a wonderful tool that helps researchers (1) save and execute code and (2) create high-quality reports. By using R Markdown, the methods for generating the results are transparent and easily shared. This simplifies the task of creating reproducible results.

For more information see
https://rmarkdown.rstudio.com/lesson-1.html

###R Programming

If at any point in running R code you encounter a function named "foo" that you do not recognize, or you want to check what the default settings are, you can type in the Console window

?foo

to call the help documentation.

### Using GitHub

The directions for using GitHub for homework assignments are given here:
http://www.princeton.edu/~mjs3/soc504_s2015/submitting_homework.shtml

The steps to using GitHub are:

1. Fork  
2. Clone
3. Commit
4. Push
5. Pull Request 

We will use use GitHub Classroom to simplify the process and keep things private. With GitHub Classroom every homework assignment will arrive as a link in an email message. When you click on this link, you will 'Fork' (step 1) the homework repository directly to your GitHub account. The link will appear on the GitHub Classroom page. Clink on the link from the Classroom page and it will open the repository in your GitHub account. From here we continue with steps 2-4 above. Step 5 will also be handled by GitHub Classroom.

If you have this on your computer, you've successfully accomplished steps 1 and 2 above. Your homework is to follow the directions below, answer any questions, and turn in the homework by following steps 3-5. 

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

r <- getOption("repos")
r["CRAN"] <- "http://cran.cnr.berkeley.edu/"
options(repos = r)

if(!require(tidyverse)){
  install.packages("tidyverse")
}
if(!require(readxl)){
  install.packages("readxl")
}

library("tidyverse")
library("readxl")
```

## 1. Edit Homework header

Edit the author slot to give your name.

## 2. Read data files from computer

In class we learned how to load data directly from GEO. Now let's learn to read data from several different formats.

I've put some example data files in the directory named /data/ExampleFiles. First, let's check our current directory.
```{r whereamI}
getwd()
```

This shows we're in the Assignments directory (NOTE: R is sensitive to capital letters). We use unix commands to move between directories. This is how we can list the files in the /data/ExampleFiles directory.

```{r filenames}
list.files("../data/ExampleFiles")
```

Great. So now let's read these files in order.

###2a. Read in *.csv file

```{r readcsv}
datadir=c("../data/ExampleFiles")
fname=paste(datadir,"/GSM2776-sample-description.csv",sep="")
fname
data=read.csv(fname,header=T,row.names=1)
head(data,n=4)
```

We can subset the matrix by rows, or by columns
```{r datasubset}
data[c("GSM277628","GSM277622"),]
data[c(4:2),c("gene","replicate")]
```

Notice how we can reorder the rows of the data matrix. This will be useful for linking gene expression data sets and covariate data from the same samples.

And we can get some tabular summaries as follows:
```{r tabulate}
table(data$gene,data$environ)
with(data,table(gene,environ))
rm(data)
```

###2b.  Read in space-delimitted *.txt  file

```{r readtable}
fname=paste(datadir,"/GSM2776-sample-description.txt",sep="")
data=read.table(fname,header=T,row.names=1)
head(data)
```

###2c.  Read in *.xls  file

```{r readxls}  
fname=paste(datadir,"/GSM2776-sample-description.xls",sep="")
data=read_xls(fname,sheet="thisone")
head(data,n=4)
rm(data)
```

## 3. Reproducible Research

The command sessionInfo() will dump information about the operating system you are running and packages you have loaded. This information, along with data and code, should allow someone to reproduce any results in the future. 
```{r sessionInfo}
sessionInfo()
```

###QUESTION: Why might a result might NOT be reproducible? 

Hint: Consider the following 2 figures.
```{r rnorm}
par(mfrow=c(1,2))
hist(rnorm(100))
hist(rnorm(100))
```

## 4. Knit and Turn in Assignment

Please turn in both your Homework1.Rmd file as well as the *.html output using GitHub. You will need to Commit the changes (a commented time/date 'save') and Push the changes from your laptop back to your GitHub homework repository. GitHub classroom will email me when these steps are completed (step 5, or Pull request). 



