---
title: "OSCA: Quick Start to single-cell RNA seq"
author: "ks"
date: "`r Sys.Date()`"
output: html_document
---

# {.tabset}

## Dataset

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
start <- Sys.time()
if (!require(scRNAseq)) BiocManager::install("scRNAseq")
if (!require(scater)) BiocManager::install("scater")
if (!require(scran)) BiocManager::install("scran")
if (!require(pheatmap)) BiocManager::install("pheatmap")
```

The following code is adapted from:
http://bioconductor.org/books/3.17/OSCA.intro/analysis-overview.html#quick-start-simple

The dataset is single cell expression from mouse retina. Begin by loading the data from the scRNAseq library & discard low quality cells. I do not plan to discuss the data cleaning steps, but you can read more on data cleaning here in Chapter 1 Quality Control (http://bioconductor.org/books/3.17/OSCA.basic/quality-control.html).

```{r loaddata-qc}
library(scRNAseq)
sce <- MacoskoRetinaData()
loaddatatime <- Sys.time()
loaddatatime - start
# Quality control (using mitochondrial genes).
library(scater)
is.mito <- grepl("^MT-", rownames(sce))
qcstats <- perCellQCMetrics(sce, subsets=list(Mito=is.mito))
filtered <- quickPerCellQC(qcstats, percent_subsets="subsets_Mito_percent")
sce <- sce[, !filtered$discard]
```


```{r data-obj}
sce
```

This object is an entire database, formatted in an object of the class: SingleCellExperiment. The expression data is in a matrix of `r nrow(sce)` genes and `r ncol(sce)` cells from a single sample.

What does the counts matrix look like?  It has lots of 0 counts, where to save space the 0 counts are not recorded. We call this a sparse matrix. Here are the data for the first 5 rows and columns (genes and cells). 
```{r counts}
counts(sce)[1:5,1:5]
```


## Dimension Reduction

Their analysis follows these steps:    
1. Select the top 10% most variable genes       
2. Run PCA on the selected genes and save the top 25 dimensions     
3. Cluster the samples using the top 25 PCs and label the samples by their identified subgroups    
4. Visualize the data: apply UMAP to the top 25 PCs and color the samples by their identified clusters.

First, we normalize the sequence counts across the cells. We'll discuss this in more detail later in the semester, but basically it's a step to remove technical variation between cells. 
```{r norm-counts}
# Normalization.
sce <- logNormCounts(sce)
logcounts(sce)[1:5,1:5]
```


```{r feat-selection}
# Feature selection
library(scran)
dec <- modelGeneVar(sce)
hvg <- getTopHVGs(dec, p = 0.1)

# PCA.
library(scater)
set.seed(1234)
sce <- runPCA(sce, ncomponents=25, subset_row=hvg)

# Clustering.
library(bluster)
colLabels(sce) <- clusterCells(sce, use.dimred='PCA',
    BLUSPARAM=NNGraphParam(cluster.fun="louvain"))    
```

The cluster labels are stored in sce$label. Here's what the data look like in PC space. The cells are colored by their cluster label.

```{r}
plotReducedDim(sce, dimred="PCA", colour_by="label")
```

We have such a large number of data points and the clusters are hard to discern in a 2-dimensional space. Let's apply the more modern techniques that use a nearest-neighbor graph approach to data visualization. 

## UMAP

```{r viz-UMAP, fig.cap = 'UMAP plot of the retina dataset, where each point is a cell and is colored by the assigned cluster identity.'}
# Visualization.
startumap <- Sys.time()
sce <- runUMAP(sce, dimred = 'PCA')
plotUMAP(sce, colour_by="label")
endumap <- Sys.time()
endumap - startumap
```


## TSNE

TSNE will take much longer to run.

```{r viz-TSNE, fig.cap = 'TSNE plot of the retina dataset, where each point is a cell and is colored by the assigned cluster identity.'}
# Visualization.
starttsne <- Sys.time()
sce <- runTSNE(sce, dimred = 'PCA')
plotTSNE(sce, colour_by="label")
endtsne <- Sys.time()
endtsne - starttsne
```

## SCE Object

Let's look at the data object with all its stored components. Each of the new data summaries has been added to the data object (logcounts, PCA, UMAP, TSNE).

```{r updated-sce-obj}
sce
```

## Find Markers

Compare each cluster to each of the remaining clusters, and identify the top marker genes that are differentially expressed between the two subgroups. 

What are the cluster sample sizes? Do we need to use a moderated t-statsitic or can we use an ordinary one?

```{r cluster-sizes}
table(sce$label)
```

These samples are big so we can use pairwise t-tests.

### Pairwise t-tests

We can filter these to just look at upregulated genes that show 2-fold higher expression in the cluster of interest.
```{r}
markers <- findMarkers(sce, test.type="t", direction="up", lfc=1) 
# plot log-fold changes for one cluster over all other clusters
# selecting cluster 1
interesting <- markers[[1]]
best_set <- interesting[interesting$Top <= 5, ]
logFCs <- getMarkerEffects(best_set)
pheatmap(logFCs, color=viridis::plasma(100))
```

The plotExpression() function allows us to see the distribution of the top genes by cluster.

```{r}
# plot log-transformed normalized expression of top genes for one cluster
top_genes <- head(rownames(interesting))

plotExpression(sce, x = "label", features = top_genes,
               color_by = "label")
```

### Wilcoxon tests

And, since these sample sizes are so large, we can switch to a non-parametric test which is preferred when looking for a marker for predicting subgroup. 

```{r marker-detection}
# Marker detection.
markers <- findMarkers(sce, test.type="wilcox", direction="up")

# returns a list with one DataFrame per cluster
markers
```

```{r}
# plot AUCs for one cluster over all other clusters
# selecting cluster 1
interesting <- markers[[1]]
best_set <- interesting[interesting$Top <= 5, ]

AUC <- getMarkerEffects(best_set,prefix = "AUC")
pheatmap(AUC, color=viridis::plasma(100))
```

```{r}
# plot log-transformed normalized expression of top genes for one cluster
top_genes_wx <- head(rownames(interesting))

plotExpression(sce, x = "label", features = top_genes_wx,
               color_by = "label")

```


## Session Info

```{r runtime}
end <- Sys.time()
end - start
```


```{r sI}
sessionInfo()
```



