---
title: "Penalized Regression"
author: "PM579"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# {.tabset}

## Background

This example is based on the code Josh Day's example from a website that no
longer exists:
https://www4.stat.ncsu.edu/~post/josh/LASSO_Ridge_Elastic_Net_-_Examples.html

There used to be a website that gives 3 simulation models for comparing ridge, lasso, and elastic net regression.

The StatQuest video uses the model in example 1 for teaching how to perform penalized regression in R using the function glmnet(). I'm going to have us run the code from the website and the video and interpret the results.

Example 1 uses the following model:

n=1000 observations

p=5000 features

All predictors are created as (standardized) normal random variates. Then it computes the observed y by adding normally distributed random error to the true regression model. 

Parameters = 15-1's, 4985- 0's

which measure the effect sizes for the p features. All predictor variables are independent.

**Generate data**

```{r generate data}
# Generate data
set.seed(42)  # Set seed for reproducibility
n <- 1000  # Number of observations
p <- 5000  # Number of predictors included in model
real_p <- 15  # Number of true predictors
x <- matrix(rnorm(n*p), nrow=n, ncol=p)
y <- apply(x[,1:real_p], 1, sum) + rnorm(n)

# Split data into train (2/3) and test (1/3) sets
train_rows <- sample(1:n, .66*n)
x.train <- x[train_rows, ]
x.test <- x[-train_rows, ]

y.train <- y[train_rows]
y.test <- y[-train_rows]
```

## Penalized Regression

```{r libs}
r <- getOption("repos")
r["CRAN"] <- "http://cran.wustl.edu/"
options(repos = r)

if (!require(glmnet)) install.packages("glmnet")
library(glmnet)
```


**Fit models**

```{r fitm}
# Fit models 
# (For plots on left):
fit.lasso <- glmnet(x.train, y.train, family="gaussian", alpha=1)
fit.ridge <- glmnet(x.train, y.train, family="gaussian", alpha=0)
fit.elnet <- glmnet(x.train, y.train, family="gaussian", alpha=.5)


# 10-fold Cross validation for each alpha = 0, 0.1, ... , 0.9, 1.0
# (For plots on Right)
for (i in 0:10) {
    assign(paste("fit", i, sep=""), cv.glmnet(x.train, y.train, type.measure="mse", 
                                              alpha=i/10,family="gaussian"))
}
```

**plot solution paths and cv mse**

```{r plots}
# Plot solution paths:
par(mfrow=c(3,2))
# For plotting options, type '?plot.glmnet' in R console
plot(fit.lasso, xvar="lambda")
plot(fit10, main="LASSO")

plot(fit.ridge, xvar="lambda")
plot(fit0, main="Ridge")

plot(fit.elnet, xvar="lambda")
plot(fit5, main="Elastic Net")
```

**mse on test set**

```{r msetestset}
yhat0 <- predict(fit0, s=fit0$lambda.1se, newx=x.test)
yhat1 <- predict(fit1, s=fit1$lambda.1se, newx=x.test)
yhat2 <- predict(fit2, s=fit2$lambda.1se, newx=x.test)
yhat3 <- predict(fit3, s=fit3$lambda.1se, newx=x.test)
yhat4 <- predict(fit4, s=fit4$lambda.1se, newx=x.test)
yhat5 <- predict(fit5, s=fit5$lambda.1se, newx=x.test)
yhat6 <- predict(fit6, s=fit6$lambda.1se, newx=x.test)
yhat7 <- predict(fit7, s=fit7$lambda.1se, newx=x.test)
yhat8 <- predict(fit8, s=fit8$lambda.1se, newx=x.test)
yhat9 <- predict(fit9, s=fit9$lambda.1se, newx=x.test)
yhat10 <- predict(fit10, s=fit10$lambda.1se, newx=x.test)

mse0 <- mean((y.test - yhat0)^2)
mse1 <- mean((y.test - yhat1)^2)
mse2 <- mean((y.test - yhat2)^2)
mse3 <- mean((y.test - yhat3)^2)
mse4 <- mean((y.test - yhat4)^2)
mse5 <- mean((y.test - yhat5)^2)
mse6 <- mean((y.test - yhat6)^2)
mse7 <- mean((y.test - yhat7)^2)
mse8 <- mean((y.test - yhat8)^2)
mse9 <- mean((y.test - yhat9)^2)
mse10 <- mean((y.test - yhat10)^2)
```

```{r printmse}
cbind.data.frame(alpha=seq(0,1,0.1),
                 mse = c(mse0,mse1,mse2,mse3,mse4,mse5,
                         mse6,mse7,mse8,mse9,mse10))
```



## StatQuest code

**Generate data**

We have the same model, but the analysis generates the results differently. 
Here's the code used in the video.
## https://github.com/StatQuest/ridge_lasso_elastic_net_demo/blob/master/ridge_lass_elastic_net_demo.R


```{r simdata}
# Generate data
set.seed(42)  # Set seed for reproducibility
n <- 1000  # Number of observations
p <- 5000  # Number of predictors included in model
real_p <- 15  # Number of true predictors
x <- matrix(rnorm(n*p), nrow=n, ncol=p)
y <- apply(x[,1:real_p], 1, sum) + rnorm(n)

# Split data into train (2/3) and test (1/3) sets
train_rows <- sample(1:n, .66*n)
x.train <- x[train_rows, ]
x.test <- x[-train_rows, ]

y.train <- y[train_rows]
y.test <- y[-train_rows]
```

**Fit models**

```{r cvglmnet}
## alpha = 0, Ridge Regression
alpha0.fit <- cv.glmnet(x.train, y.train, type.measure="mse",
                        alpha = 0, family="gaussian")
## now let's run the Testing dataset on the model created for
## alpha = 0 (i.e. Ridge Regression).
alpha0.predicted <- predict(alpha0.fit, s = alpha0.fit$lambda.1se,
                            newx = x.test)
## Lastly, let's calculate the Mean Squared Error (MSE) for the model
## created for alpha = 0.
## The MSE is the mean of the sum of the squared difference between 
## the predicted 'y' values and the true 'y' values in the 
## Testing dataset...
mean((y.test - alpha0.predicted)^2)


## alpha = 1, Lasso Regression
alpha1.fit <- cv.glmnet(x.train, y.train, type.measure="mse",
                        alpha = 1, family="gaussian")
alpha1.predicted <- predict(alpha1.fit, s = alpha1.fit$lambda.1se,
                            newx = x.test)
mean((y.test - alpha1.predicted)^2)

## alpha = 0.5, a 50/50 mixture of Ridge and Lasso Regression
alpha0.5.fit <- cv.glmnet(x.train, y.train, type.measure="mse",
                        alpha = 0.5, family="gaussian")
alpha0.5.predicted <- predict(alpha0.5.fit, 
                              s = alpha0.5.fit$lambda.1se,
                            newx = x.test)
mean((y.test - alpha0.5.predicted)^2)
```

**Pick alpha by min mse**

```{r fit-models}
list.of.fits <- list()
for (i in 0:10){
  fit.name <- paste0("alpha", i/10)
  
  list.of.fits[[fit.name]] <-
    cv.glmnet(x.train, y.train, type.measure = "mse", alpha = i/10,
              family = "gaussian")
}

## Now we see which alpha (0, 0.1, ... , 0.9, 1) does the best job
## predicting the values in the Testing dataset.
results <- data.frame()
for (i in 0:10){
  fit.name <- paste0("alpha", i/10)

  ## Use each model to predict 'y' given the Testing dataset
  predicted <-
    predict(list.of.fits[[fit.name]] ,
           s = list.of.fits[[fit.name]]$lambda.1se, newx = x.test)
  
  ## Calculate the Mean Squared Error...
  mse <- mean((y.test-predicted)^2)
  
  ## Store the results
  temp  <- data.frame(alpha = i/10, mse = mse, fit.name = fit.name)
  results <- rbind(results,temp)
}
## View the results
results
```

How do these compare to the results from the video?

https://www.youtube.com/watch?v=ctmNq7FgbvI&t=203s

(16:41 min)

Why do you think the results differ despite using the same random number seed? 

## Session Information

```{r sI}
sessionInfo()
```
