---
title: "ChatGPT activity 2: Test size under Null"
author: "ks"
date: "`r Sys.Date()`"
output: html_document
---

# {.tabset}

## Simulation set up 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(limma)
library(tidyverse)
require(MASS)
```

Let's estimate the distribution of standard deviations from our real data set.  

Step 1: Subset the (real) data. 

Step 2: Analyze. 

This will be for testing differential expression over time (day) in three lungs.
```{r aecdat}
load(file = c("../data/aec.rda"))

subEList <- function(ELst,ids){
   sELst <-NULL
     sELst$targets <- ELst$targets[ids,]
     sELst$E       <- ELst$E[,ids]
     sELst$genes   <- ELst$genes
     sELst
}
idx <- which(aec$targets$day %in% c("D0","D2"))
aecdat <- subEList(aec,idx)
table(aecdat$targets$day)
```
  
I will throw away one D0 technical replicate because it would throw off the variance estimates in my study of biological samples.  Then I estimate the biological sds from the subset of aec data, using the empirical Bayes estimates from the limma package. I use this distribution of sds for my simulation study.

```{r subset}  
aecsub <- NULL
aecsub$targets <-aecdat$targets[!duplicated(aecdat$targets[,c("age","day")]),]
aecsub$E <- aecdat$E[,rownames(aecsub$targets)]
table(aecsub$targets$day)

design <- model.matrix(~ factor(day) + factor(age), data = aecsub$targets)
fit  <- limma::lmFit(aecsub$E,design)
efit <- limma::eBayes(fit)
sds <- sqrt(efit$s2.post)
```


I will simulate data the following default design settings:   

* 6 samples

* 10,000 genes:  0 differentially expressed 

```{r design-vals}
nsamp <- 6       # number of samples
n1     <- nsamp/2    # equal sample sizes in each group
ng     <- 10000     # number of genes
set.seed(99)
ssds <- sample(sds,ng) # sample, and randomize the order
```

I will simulate data under 2 scenarios. The first will simulate data from the null (without loss of generality, mean=0), for 6 independent samples. The second will do it for paired samples, i.e., from the same lung.

## Scenario 1: Indep

Function to simulate data under the null for the independence model.

```{r simdat}
simdat <- function(se,nsamp=6,ng=10000,mn=0){
  n1 <- nsamp/2   
  gp  <- factor(rep(c("group1", "group2"), each=n1))
    sdat <- replicate(nsamp,rnorm(n = ng, mean = mn, sd = se))
  list(E=sdat,targets=gp)
}
```

Function to simulate the data, analyze it, and compute the number of false positives.

```{r simdat-analyze}
sim.indep <- function() {
                    sim1 <- simdat(ssds)

                    design <- model.matrix(~ factor(sim1$targets))
                    fit  <- limma::lmFit(sim1$E,design)
                    efit <- limma::eBayes(fit)

                    fpr<- mean(efit$p.value<0.05)
                    fpr
}
```

Estimate the false-positive rate for 1 sample. 

```{r indep1}
set.seed(88)
sim.indep()
```

Now let's do it for 10,000 samples and take the average.

What's an easy way to repeat the function multiple times and save the output? Let's ask ChatGPT.


```{r sim5xs}
set.seed(123)

```


Run this another way to check that the results are correct.
```{r}

```

Suppose I want to make sure this reflects more studies. Then, I might want to do this 10,000 times. If I run the code above replacing 5 with 10000, it takes too long. Let's ask ChatGPT for code that runs faster.


```{r parallel-sim10000xs}


```


```{r parallel-sim10000xs1}

```

```{r parallel-sim10000xs2}

```

Results? Conclusions?


## Scenario 2: subj-to-subj var

My data are not all independent. I know this from the design. I have time course data for cells from 3 lungs. We expect there will be correlation between measurements from the same lung. How can I simulate that?  We often specify a small variation in the average measure between subject (here lung). 

Now I go back and create subject level correlation between measures. 

```{r sim2}

simdat.subj <- function(se,nsamp=6,ng=10000,mn=0){
  n1 <- nsamp/2   
  gp  <- factor(rep(c("group1", "group2"), each=n1))
  subj <- rep(paste0("subj",1:n1),each=2)
     sdat <- replicate(nsamp,rnorm(n = ng, mean = mn, sd = se))
  sdat[,3:4] <- sdat[,3:4] + 0.10
  sdat[,5:6] <- sdat[,5:6] - 0.05 
  #sdat[,3:4] <- sdat[,3:4] + rnorm(ng,mean = 0.10, sd = median(se))
  #sdat[,5:6] <- sdat[,5:6] - rnorm(ng,mean = -0.05, sd = median(se))  
  list(E=sdat,targets=cbind.data.frame(gp,subj))
}
```


It will be interesting to compare the results for 2 models.  In the first model, I treat all samples as independent (2 sample t-test). In the second, I want to adjust for subject, which is equivalent to doing a paired t-test.

Here's the function to simulate 1 data set and analyze it using the two models above.

```{r rsim2}
sim.paired <- function() {
      sim2 <- simdat.subj(ssds)

      # ignore subject effect
      design1 <- model.matrix(~ factor(gp),data = sim2$targets)
      fit1  <- limma::lmFit(sim2$E,design1)
      efit1 <- limma::eBayes(fit1)
      fdr1 <- mean(efit1$p.value[,2]<0.05)

      # adjust for subject
      design2 <- model.matrix(~ factor(gp) + factor(subj),data = sim2$targets)
      fit2  <- limma::lmFit(sim2$E,design2)
      efit2 <- limma::eBayes(fit2)
      fdr2 <- mean(efit2$p.value[,2]<0.05)
   
      list(fdr1=fdr1, fdr2=fdr2)
}
```

Let's estimate the false-positive rates from 1 data set.

```{r sim2-fpr}
set.seed(100)  
sim.paired()
```

And run it for more 

```{r}
set.seed(100)
replicate(2, sim.paired())
```


Let's run this 10,000 times to estimate the false-positive rate.

```{r}
start_time <- Sys.time()
results <- 
Sys.time() - start_time

paste(" indep model FP rate:", round(mean(unlist(results[1,])),4))
paste("paired model FP rate:", round(mean(unlist(results[2,])),4))
```


Now, I want to see what happens when we analyze the data using a linear mixed effects model to account for correlation instead of a fixed effect model (what we did above).

```{r lmefnc}
lmefnc <- function(){
    sim2 <- simdat.subj(ssds)
    
    design <- model.matrix(~gp , data = sim2$targets)
    #Then we estimate the correlation between measurements made on the same subject:
    corfit <- duplicateCorrelation(sim2$E,design,block=factor(sim2$targets$subj))
    #corfit$consensus
    #Then this inter-subject correlation is input into the linear model fit:
     fit <- lmFit(sim2$E,design,
                     block=sim2$targets$subj,
                     correlation=corfit$consensus)

    # fit has 2 coefficients: intercept + Coef of interest.  Let's just save the test results for     the second coefficient, the coefficient of interest. 
    contr.matrix <- as.matrix(c(0,1))
    colnames(contr.matrix) <- c("day2")
    fit <- contrasts.fit(fit,contr.matrix)
    efit <- limma::eBayes(fit, trend = TRUE)
    
    fpr <- mean(efit$p.value < 0.05)
    fpr
}

set.seed(100)
start_time <- Sys.time()
lmefnc()
Sys.time() - start_time
```
At almost 5 seconds/run, let's only run 100 of these to estimate the false-positive rate. 


```{r}
start_time <- Sys.time()
results <- 
Sys.time() - start_time

paste(" lme model FP rate:", round(mean(results),4))
```

I can imagine why this might be an exaggeration of the effect. Any ideas?

Instead of shift, try:   
- N(shift,median(sds)), or 
- N(0,median(sds))
 
## SessionInfo

```{r sessioninfo}
sessionInfo()
```
