---
title: "ChatGPT Activity 1"
author: "ks"
date: "`r Sys.Date()`"
output: html_document
---

# {.tabset}

## Activities

1. Subset the homework 3 data (study of lung cell differentiation) to day 0 and day 2   
2. Perform a differential expression analysis comparing the data at days 0 and 2 
3. Generate a figure to summarize the results 
4. Describe the results

**Assignment**: Using the same lung data set from homework 2, you will perform a differential expression analysis between days 0 and 2.  ...


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadlibraries}
library(bookdown)
library(matrixStats)
library(ggplot2)
library(limma)
library(dplyr)
library(tidyr)
```


## Data 

Load the human alveolar epithelial cell dataset and subset it to days 0 and 2.

```{r aecdata}
   load("../data/aec.rda")
# write a function to subset the list objects using the column and row idx

subEList <- function(ELst,ids){
   sELst <-NULL
     sELst$targets <- ELst$targets[ids,]
     sELst$E       <- ELst$E[,ids]
     sELst$genes   <- ELst$genes
     sELst
}

   idx <- which(aec$targets$day %in% c("D0","D2"))
   aecdat <- subEList(aec,idx)
   aecdat$targets
```

These data are from Marconett et al. Integrated transcriptomic and epigenomic analysis of primary human lung cell differentiation. PLoS Genet 2013. PMC3688557.

```{r}
table(aecdat$targets$day)
```


## Differential Expression

Create a figure to display the results from the differential expression analysis. Write a figure caption that names the data display and interpret the axes.  Describe any patterns that you see in the figure.

```{r fig1,fig.cap='Volcano plot of differential expression of day 2 (n=3) and day 0 (n=4) '}
design <- model.matrix(~day, data=aecdat$targets)
#design
fit <- limma::lmFit(aecdat$E,design)
# fit has 2 coefficients: intercept + Coef of interest.  Let's just save the test results for the second coefficient, the coefficient of interest. 
contr.matrix <- as.matrix(c(0,1))
colnames(contr.matrix) <- c("day2")
# This will reduce the model output to the coef of interest only
fit <- contrasts.fit(fit,contr.matrix)
efit <- limma::eBayes(fit)

limma::volcanoplot(efit,  style = "p-value", highlight = 10, names = aecdat$genes$Symbol, hl.col="blue",
            xlab = "Log2 Fold Change", ylab = NULL, pch=16, cex=0.35)
```

Report the number of results that are significant at a 5% false-discovery rate. Then create a table of the top 10 differentially expressed features.

```{r decideTestsv}
summary(de <- limma::decideTests(efit, p=0.05))
```

```{r TopT}
options(digits=3)
topT <- limma::topTable(efit,n=10)
topT
```

Here is the annotation for these features:

```{r genenames}
aecdat$genes[rownames(topT),]
```

NOTE: Some features do not have Entrez Gene IDs. 

```{r}
sum(is.na(aecdat$genes$Entrez_Gene_ID))
```

## Analysis concerns

Concern 1. Day 0 has a technical replicate      
Concern 2. Time series measures are matched/paired   

Regarding 1:   

Analyzing Day 0 as having 4 independent measurements is wrong because it assumes we have more information than we have. We have 2 options, averaging the 2 technical replicates or picking 1 of them. The average will have a different variance than a single measurement, so for simplicity, I will pick 1 observation and remove the other. 

Ask ChatGPT for help removing the row with duplicate day/age. Then go remove that sample from the expression data too.

```{r rmduplicate}

```


Check the results of data filtering.

```{r}
aecdat$targets
identical(rownames(aecdat$targets),colnames(aecdat$E))
```



## Diff Expr 3v3

Perform differential expression on filtered data set of n3=3 vs n=3.

```{r fig2,fig.cap='Volcano plot of differential expression of day 2 (n=3) and day 0 (n=3)  after removing technical replicate'}
design <- model.matrix(~day, data = aecsub$targets)
#design
fit <- limma::lmFit(aecsub$E,design)
# fit has 2 coefficients: intercept + Coef of interest.  Let's just save the test results for the second coefficient, the coefficient of interest. 
contr.matrix <- as.matrix(c(0,1))
colnames(contr.matrix) <- c("day2")
# This will reduce the model output to the coef of interest only
fit <- contrasts.fit(fit,contr.matrix)
efit <- limma::eBayes(fit)

limma::volcanoplot(efit,  style = "p-value", highlight = 10, names = aecsub$genes$Symbol, hl.col="blue",
            xlab = "Log2 Fold Change", ylab = NULL, pch=16, cex=0.35)
```

```{r decideTestsv2}
summary(de2 <- limma::decideTests(efit, p=0.05))
```

Let's compare this to the analysis of 7 samples.

```{r}
table(de,de2)
```

Over 1000 fewer features are differentially expressed, which makes sense because previously we exaggerated the accuracy of our measures by including a technical replicate.

## repeat, adj for subject using age

Now I should have accounted for the paired nature of our measurements. Let's do that now and see how much changes if I adjust for patient lung, to look at within-subject differences over time. (paired test). Let's ask ChatGPT how to adjust the model for lung age (subject). 

ChatGPT prompt:

```{r fig3,fig.cap='Volcano plot of differential expression of day 2 (n=3) vs day 0 (n=3), adjusted for subject'}


```


Number of differentially expressed features (FDR-adjusted p<0.05):
```{r decideTestsv3}
summary(de3 <- limma::decideTests(efit, p=0.05))
```

Now compare this to the analysis of 3 vs 3 without adjustment (2-sample t-test).

```{r}
table(de2,de3)
```

I find ___________(more/fewer) differentially expressed genes using the age-adjusted model.  Why?


Is that the best way to adjust for subject?  Let's look at the design matrix:
```{r}
design
```

Since subject is a nuisance variable, we should try to adjust for it in the most flexible manner possible. Is that the most flexible model to adjust for subject/lung effects?  If no, how would you improve the model?


```{r fig4,fig.cap='Volcano plot of differential expression of day 2 (n=3) vs day 0 (n=3), adjusted for subject'}


```


```{r decideTestsv4}
summary(de4 <- limma::decideTests(efit, p=0.05))
```

Compare results from the previous adjusted model to the new one and interpret the results.
```{r}
table(de3,de4)
```

Does this make sense?


There is a third way one might consider modeling the correlation. That is to model the subject-level variation using a random effect where each lung has its own average expression level. This puts a parametric distribution on the variation in expression between lungs. How would that affect our results?

## Model correlation

```{r fig5,fig.cap='Volcano plot of differential expression of day 2 (n=3) vs day 0 (n=3), using linear-mixed effects model'}
design <- model.matrix(~ day, data = aecsub$targets)
#Then we estimate the correlation between measurements made on the same subject:
corfit <- duplicateCorrelation(aecsub$E,design,block=factor(aecsub$targets$age))
corfit$consensus
#Then this inter-subject correlation is input into the linear model fit:
fit <- lmFit(aecsub$E,design,
             block=aecsub$targets$age,
             correlation=corfit$consensus)

contr.matrix <- as.matrix(c(0,1))
colnames(contr.matrix) <- c("day2")

fit <- contrasts.fit(fit,contr.matrix)
efit <- limma::eBayes(fit, trend = TRUE)

decorr  <- decideTests(efit, p.value = 0.05)
summary(decorr)
```


```{r decideTestsvc5}
table(de4,decorr)
```

It finds _______________(more/fewer) features.  Is that good or bad?  To answer that, I would want to run
a simulation and estimate the false-positive rate of the different models. Do any of the models have an inflated error rate?

I want to run a simulation to know if I should model the correlation with a random effect, or adjust for it using a fixed effect.  We can pursue that question in: w09-chatgpt-activity2.Rmd

## Report computing environment

Now output the computing environment listing the different software versions installed on your computer.

```{r sessioninfo}
  sessionInfo()
```
