---
title: "Numbered Figure Captions"
author: "K Siegmund"
date: today
format: html
execute:
  echo: true
  warning: false
---

```{r}
#| label: rlibraries
#| echo: false
library(ggplot2)
```

```{r}
#| label: loaddata
load("../data/JBC 2012/jbcdat.rda")
dim(jbcdat$E)
```

# Principal Components Analysis

```{r}
#| label: fig-pca
#| fig-cap: "PCA of 24 gene expression arrays."
tE <- t(jbcdat$E)

my.pca <- prcomp(tE, center = TRUE,
                 scale = matrixStats::colMads(tE),
                 retx = TRUE)
dfx <- as.data.frame(my.pca$x)

ggplot(dfx, aes(x = PC1, y = PC2, color = jbcdat$targets$type)) +
  geom_point(size = 2.5) +
  theme(text = element_text(size = 20),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12)) +
  labs(color = "Treatment")
```

Now let's repeat the analysis on just the top 500 variable genes. I will use median absolute deviation for measuring variability.

```{r}
#| label: fig-pca2
#| fig-cap: "PCA of 24 gene expression arrays (500 most variable genes)."
fmad <- matrixStats::rowMads(jbcdat$E)
fidx <- order(fmad, decreasing = TRUE)[1:500]

subsetE <- tE[, fidx]

my.pca <- prcomp(subsetE, center = TRUE,
                 scale = matrixStats::colMads(subsetE),
                 retx = TRUE)
dfx <- as.data.frame(my.pca$x)

ggplot(dfx, aes(x = PC1, y = PC2, color = jbcdat$targets$type)) +
  geom_point(size = 2) +
  labs(color = "Treatment")
```

Why might this show better clustering?

# Multi-dimensional Scaling

```{r}
#| label: fig-mds
#| fig-cap: "MDS plot of 24 gene expression arrays (500 features, see methods for details)."
limma::plotMDS(jbcdat$E, pch = 16,
               col = unclass(jbcdat$targets$type),
               xlim = c(-1.5, 1.5), ylim = c(-1, 1),
               main = "MDS plot")
legend(-1.5, 1, levels(jbcdat$targets$type),
       pch = 16, col = order(levels(jbcdat$targets$type)),
       cex = 0.75)
```

This result resembles the PCA result, but it can be different due to the different variable selection method. MDS is based on a pairwise dissimilarity matrix of the samples, with separate feature filtering performed by default for each pair of samples.

# Session info

```{r}
#| label: session-info
sessionInfo()
```
